\chapter{Rheology}

The reader is referred to Barnes \cite{barn99}
for a discussion and review of non-linear viscous rheologies and 
to Coussot \cite{cous14} for a review of experimental data for yield stress fluid
flows. See also Tanner \& Tanner \cite{tata03} for a summary of Heinrich Hencky's 
scientific work on rheology. 

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Relevant literature}

\begin{itemize}
\item Plasticity and Geomechanics, Davis and Selvadurai. \cite{dase02}
\item Elasticity and Geomechanics, Davis and Selvadurai. \cite{dase96}
\item Rheology of the Earth, Ranalli. \cite{ranalli}
\item Deformation of Earth materials, Karato. \cite{kara08}
\item Fundamentals of the Theory of Plasticity, Kachanov. \cite{kacha04} 
\item Computational methods for plasticity, de Souza Neto et al. \cite{depo}
\item Computer simulation of dynamic phenomena, M. wilkins \cite{wilk}
\item Continuum theory of plasticity, Khan and Huang \cite{khhu}
\item Theory of plasticity, Chakrabarty \cite{chakrabarty}
\item Zienkiewicz Taylor \cite{zita2}
\item Rheology Principles, Macosko \cite{macosko}
\item Computational Inelasticity, Simo and Hughes \cite{simohughes} 
\item Lectures on Visco-Plastic Fluid Mechanics, G. Ovarlez \& S. Hormozi \cite{ovho19}
\item Complex fluids, P. Saramito \cite{saramito}
\end{itemize}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Linear viscous aka Newtonian} 
\input{newtonian_fluids}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Power-law model \label{ss:powerlaw}} 
\input{powerlaw}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Carreau model}
\input{carreau_model}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Bingham model} \label{sec:bingham}
\input{bingham_model}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Herschel-Bulkley visco-plastic model}
\input{herschel_bulkley_model}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Casson model}
\input{casson_model}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Ellis model\label{ss:ellis}}
\input{ellis_model}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{One model to rule them all? \label{ss:cross}}
\input{rule_them_all}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Dislocation and Diffusion creep}
\input{dislocation_diffusion_creeps}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The von Mises failure criterion}\label{sec:vMcriterion}
\input{vMcriterion}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Tresca failure criterion}\label{sec:trcriterion}
\input{trcriterion}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Mohr-Coulomb failure criterion}\label{sec:mccriterion}
\input{mccriterion}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Drucker-Prager failure criterion \label{sec:dpcriterion}}
\input{dpcriterion}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Griffith-Murrell failure criterion}
\input{griffith_murrell}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Cam-clay failure criterion}
\input{camclay}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The failure envelope, or yield surface \label{ss:envelope}}
\input{yield_surface}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Peierls creep}
\input{peierls}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Stress limiting (pseudo-plastic) rheology}
\input{pseudo_plastic_rheology}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Arrhenius law} 
\input{arrhenius_law}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Frank-Kamenetskii rheology}
\input{frank_kamenetskii}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Glen's law for ice}\label{ss:glen}
\input{glen_flow_law}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Strain rate partitioning across deformation mechanisms}\label{ss:srpart}
\input{strain_rate_partitioning}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Anisotropic viscosity}
\input{anisotropic_viscosity}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Rheology of the lithosphere}
\input{lithosphere_rheology}

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Perzyna model}\label{sec:perzyna}
\index{general}{Perzyna Model}
\input{perzyna}

\input{perzyna_vM}

\input{perzyna_chpe15}

\input{perzyna_others}






