\begin{flushright} {\tiny {\color{gray} fsbc\_annulus.tex}} \end{flushright}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

In the context of geodynamical modelling we often wish to prescribed free-slip 
boundary conditions on a given boundary of the domain. If the domain is a rectangle
which sides align with the Cartesian axis, then fixing $\upnu_x=0$ or $\upnu_y=0$
is simple and does indeed insure free-slip boundary conditions. 

However the situation is much more complicated in the case of a curved boundary, 
such as for instance the inner and outer boundaries of an annulus or spherical shell.

If the curved boundary is a circular, the procedure is as follows:
\begin{enumerate}
\item identify the node on the boundary which is to be fixed. 
\item compute its coordinate angle $\theta$ (and $\phi$ in 3D) 
\item do a rotation so as to bring it back onto the x-axis (2D) or z-axis (3D)
\item apply free slip boundary condition (now easy since parallel or perpendicular to axis)
\item rotate back
\end{enumerate}

This technique is implemented in \stone 33 and \stone 96.

\paragraph{A few remarks about rotation matrices} 
In a given plane, the counter-clockwise rotation matrix by and angle $\theta$ is defined by 
\[
{\cal R}=
\left(
\begin{array}{cc}
\cos\theta & \sin\theta \\
-\sin\theta & \cos\theta
\end{array}
\right)
\]
The image of vector $\vec{V}$ by a rotation of angle $\theta$ is given by
\[
\vec{V}'={\cal R}\cdot \vec{V}
\]

Coordinate transformations of second-rank tensors involve the very same   
matrix as vector transforms. A transformation of the stress tensor ${\bm \sigma}$ ,
from the reference $xy$-coordinate system to ${\bm \sigma}'$ in a new $x'y'-$system is done as follows:
\[
{\bm \sigma}'={\cal R}\cdot {\bm \sigma}\cdot{\cal R}^T
\]



[from Wikipedia] A basic rotation (also called elemental rotation) is a rotation about one of the axes of a Coordinate system. 
The following three basic rotation matrices rotate vectors by an angle $\alpha$ 
about the x-, y-, or z-axis, in three dimensions, using the right-hand rule which codifies their 
alternating signs. 

\[
{\cal R}_x(\alpha)=
\left(
\begin{array}{ccc}
1 & 0 & 0 \\
0 & \cos\alpha & -\sin\alpha \\
0 & \sin\alpha & \cos\alpha
\end{array}
\right)
\]

\[
{\cal R}_y(\alpha)=
\left(
\begin{array}{ccc}
\cos\alpha & 0 & \sin\alpha \\
0 & 1 & 0 \\
-\sin\alpha & 0 &\cos\alpha
\end{array}
\right)
\]

\[
{\cal R}_z(\alpha)=
\left(
\begin{array}{ccc}
\cos\alpha & -\sin\alpha & 0\\
\sin\alpha & \cos\alpha & 0 \\
0 & 0 & 1 
\end{array}
\right)
\]

In my \elefant code
I first rotate around the $z$ axis by and angle $-\phi$ and then 
around axis $y$ by an angle $-\theta$ in the case of a spherical shell.

\[
{\cal R}_y(-\theta)=
\left(
\begin{array}{ccc}
\cos(-\theta) & 0 & \sin(-\theta) \\
0 & 1 & 0 \\
-\sin(-\theta) & 0 &\cos(-\theta)
\end{array}
\right)
=
\left(
\begin{array}{ccc}
\cos\theta & 0 & -\sin\theta \\
0 & 1 & 0 \\
\sin\theta & 0 &\cos\theta
\end{array}
\right)
\]

\[
{\cal R}_z(-\phi)
=
\left(
\begin{array}{ccc}
\cos(-\phi)& -\sin(-\phi) & 0\\
\sin(-\phi)& \cos(-\phi) & 0 \\
0 & 0 & 1 
\end{array}
\right)
=
\left(
\begin{array}{ccc}
\cos\phi& \sin\phi & 0\\
-\sin\phi& \cos\phi & 0 \\
0 & 0 & 1 
\end{array}
\right)
\]

These are the {\tt Rott} and {\tt Rotp} matrices in the routines.



