
\begin{flushright} {\tiny {\color{gray} howtobc.tex}} \end{flushright}

Let us consider a quadrilateral element with one degree of freedom per node and let us assume that we are solving the temperature equation. The local matrix and right-hand side vector are given by 
\[
A_{el}(4\times 4) \quad\quad {\text and} \quad\quad B_{el}(4)
\]
Let us assume that we want to impose $\tilde{T}=10$ on the third node (local coordinates numbering). For instance, having built $A_{el}$ and $B_{el}$, the system looks like :
\[
\left(
\begin{array}{cccc}
3 & 1 & 6  & 9 \\
5 & 2 & 2  & 8 \\
7 & 4 & 11 & 2 \\
9 & 6 & 4  & 3
\end{array}
\right)
\left(
\begin{array}{c}
T_1 \\ T_2 \\ T_3 \\ T_4
\end{array}
\right)
=
\left(
\begin{array}{c}
4 \\ 3 \\ 1 \\ 2
\end{array}
\right)
\]



which can be rewritten 
\[
3 T_1 + T_2 + 6 T_3 + 9 T_4 = 4
\]
\[
5 T_1 + 2T_2 + 2 T_3 + 8 T_4 = 3
\]
\[
7 T_1 + 4T_2 + 11 T_3 + 2 T_4 = 1
\]
\[
9 T_1 + 6T_2 + 4 T_3 + 3 T_4 = 2
\]
or, 
\[
3 T_1 + T_2 + \quad  + 9 T_4 = 4 - 6T_3
\]
\[
5 T_1 + 2T_2 + \quad + 8 T_4 = 3 - 2T_3
\]
\[
7 T_1 + 4T_2 + 11T_3 + 2 T_4 = 1 
\]
\[
9 T_1 + 6T_2 + \quad + 3 T_4 = 2 - 4T_3
\]


\begin{itemize}

\item \underline{Technique 1:} Replace the hereabove system by
\[
\left(
\begin{array}{cccc}
3 & 1 & 6  & 9 \\
5 & 2 & 2  & 8 \\
7 & 4 & 11 +  10^{12} & 2 \\
9 & 6 & 4  & 3
\end{array}
\right)
\left(
\begin{array}{c}
T_1 \\ T_2 \\ T_3 \\ T_4
\end{array}
\right)
=
\left(
\begin{array}{c}
4 \\ 3 \\ \tilde{T}\times (11 + 10^{12}) \\ 2
\end{array}
\right)
\]




\item \underline{Technique 2:} One can choose not to solve for $T_3$ anymore, i.e. not to consider it as a degree of freedom and therefore write:

\[
3 T_1 + T_2 + 9 T_4 = 4 - 6T_3
\]
\[
5 T_1 + 2T_2 + 8 T_4 = 3 - 2T_3
\]
\[
9 T_1 + 6T_2 +  3 T_4 = 2 - 4T_3
\]


\item \underline{Technique 3:} Since we want to impose $T_3=10$, then we can write 
\[
3 T_1 + T_2 + \quad  + 9 T_4 = 4 - 6T_3
\]
\[
5 T_1 + 2T_2 + \quad + 8 T_4 = 3 - 2T_3
\]
\[
0 + 0 + T_3 + 0 = 10
\]
\[
9 T_1 + 6T_2 + \quad + 3 T_4 = 2 - 4T_3
\]
and in matrix form :
\[
\left(
\begin{array}{cccc}
3 & 1 & 0  & 9 \\
5 & 2 & 0  & 8 \\
0 & 0 & 1 & 0 \\
9 & 6 & 0  & 3
\end{array}
\right)
\left(
\begin{array}{c}
T_1 \\ T_2 \\ T_3 \\ T_4
\end{array}
\right)
=
\left(
\begin{array}{c}
4 - A_{13} T_3\\ 3 - A_{23}T_3 \\ 10 \\ 2-A_{43} T_3
\end{array}
\right)
\]

\end{itemize}

The first technique is not a good idea in practice as it introduces very large 
values and will likely derail the solver. The second option is somewhat difficult
to implement as it means that elemental matrix and rhs sizes will change from 
element to element and it therefore requires more book-keeping.
The third technique is the one adopted throughout this document. 

As shown in \textcite{wuxl08} (2008), it is better to replace the 1 on the diagonal 
by the former diagonal term as it reduces the condition number of the matrix. 
The rhs must then be modified accordingly.

\Literature Behr (2004) \cite{behr04}

\todo{ASK DAVE for permission} This is an excerpt of an email sent to me by Dave May in May 2014: 
{\it 
Never ever ever impose bc's using a penalty approach.
For problems with a fixed mesh topology and time dependent Dirichlet domain (e.g. the segment 
of the boundary with Dirichlet bc's 
maybe change size/shape over time - for example with a true stick/slip type interface), it's nice to define the matrix 
with the dimension associated with the mesh+basis and leave all bc's in the operator. 
Leaving the bc's in the operator can be implemented in a manner which still retains the operators symmetry (assuming 
it was symmetric to begin with). This leaves the choice of what to stick on the diagonal. Simply using "1" could 
screw up the spectrum of the matrix and kill the iterative solver performance. A better choice would be to insert 
a diagonal entry closely related to the operator; e.g. something that looks like the diagonal entry 
of $\int 2 \eta \epsilon(u) : \epsilon(v) dV$ (for the discrete stress tensor term). 

Removing Dirichlet bc's entirely for the discrete operator sounds attractive. The code will like the FE theory 
and you will only be solving for variables which are "unknowns" (compared with the above). 
However, introducing a time dependent Dirichlet domain means the matrix must be re-sized, as should its non-zero structure 
be re-allocated. Also, implemented multi-grid is annoying when the Dirichlet entries are removed. In fact, most of  
the code associated with stripping out Diriclet bc's is annoying and ugly. However, removing the bcs ensures symmetry, 
it ensures the discrete operator will have a nice spectrum (c.f. the above option). Also, stripping out bcs usually 
increases overall storage as you have one representation of the discrete vectors given to the solver which will be  
of size (N-n) and in your mesh you will have a repsentation of length N. ``N'' being the total number of dofs in your 
system, ``n'' being the number of Dirichlet constrained dofs in your system. }
 
