

The success of a mixed finite element formulation crucially depends on a proper choice of the local interpolations of the velocity and the pressure. 

%........................................................................................
\subsubsection{The compatibility condition (or LBB condition, or inf-sup condition)} \label{ss:LBBcond}
\index{general}{LBB} \index{general}{Optimal Rate}

\input{lbb}


%........................................................................................
\subsubsection{Families}
\index{general}{Taylor-Hood}

The family of {\color{olive} Taylor-Hood} finite element spaces on triangular/tetrahedral 
grids is given by $P_k \times P_{k-1}$ with $k\geq 2$, 
and on quadrilateral/hexahedral grids by $Q_k \times Q_{k-1}$ with $k\geq 2$.
This means that the pressure is then approximated by continuous functions. 

These finite elements are very popular, in particular the pairs for $k=2$, i.e.
$Q_2\times Q_1$ and $P_2\times P_1$.
The reason why $k\geq 2$ comes from the fact that the 
$Q_1 \times Q_0$ (i.e. $Q_1 \times P_0$) and $P_1\times P_0$
are not stable elements (they are not inf-sup stable), as
shown in John \cite[p64]{john16} and \cite[p67]{john16}. 

\begin{remark}
Note that a similar element to $Q_2 \times Q_1$ has been proposed
and used succesfully used \cite{taho73,hota74}: it is denoted by $Q_2^{(8)} \times Q_1$ 
since the center node ('$x^2y^2$') and its associated degrees of freedom have been removed. It 
has also been proved to be LBB stable. These are also called {\color{olive} Serendipity} elements. 
\end{remark}

The {\color{olive} Raviart-Thomas} family... 

Check Scott \& Vogelius (1985) \cite{scvo85} element, see John \cite[p70]{john16}.

%........................................................................................
\subsubsection{The bi/tri-linear velocity - constant pressure element ($Q_1\times P_0$)}
\label{ss:pairq1p0}
\input{pair_q1p0}

%----------------------------------------------------------------------
\subsubsection{The bi/tri-quadratic velocity - bi/tri-linear pressure element ($Q_2 \times Q_1$)}
\label{ss:pairq2q1}
\input{pair_q2q1}

%----------------------------------------------------------------------
\subsubsection{The bi/tri-quadratic velocity - discontinuous linear pressure element ($Q_2 \times P_{-1}$)}
\label{ss:pairq2pm1}
\input{pair_q2pm1}

%----------------------------------------------------------------------
\subsubsection{The stabilised bi/tri-linear velocity - constant pressure element ($Q_1\times P_0$-stab)}
\label{ss:pairq1p0stab}
\input{pair_q1p0stab}


%----------------------------------------------------------------------
\subsubsection{The stabilised bi/tri-linear velocity -  bi/tri-linear pressure element ($Q_1\times Q_1$-stab)}
\label{ss:pairq1q1stab}
\input{pair_q1q1stab}

%----------------------------------------------------------------------
\subsubsection{The MINI triangular element ($P_1^+\times P_1$) in 2D}
\label{pair:mini}
\input{pair_mini}

%----------------------------------------------------------------------
\subsubsection{The quadratic velocity - linear pressure triangle ($P_2\times P_1$)}
\label{ss:p2p1}
\input{pair_p2p1}



%----------------------------------------------------------------------
\subsubsection{The Crouzeix-Raviart triangle ($P_2^+\times P_{-1}$)}
\label{sec:crouzeix-raviart}

Since the $P_2\times P_{-1}$ pair is not LBB stable \cite[p179]{reddybook2}, 
it is enhanced by a cubic bubble and is therefore called $P_2^+\times P_{-1}$. 

This element was first introduced in \cite{crra73}.
It is the element used in the MILAMIN code \cite{daks08}.
It is a seven-node triangle with quadratic velocity shape 
functions enhanced by a cubic bubble function and discontinuous linear interpolation for 
the pressure field \cite{cuss86}. 
This element is LBB stable and no additional stabilization techniques are required\cite{elsw}.
The '+' in its name stands for the bubble while the '-' stands for the discontinuous
character of the pressure field: once again, it is $P_1$ over the element, but discontinuous
across element edges.

\begin{remark}
Cuvelier \etal, 1986 \cite{cuss86} recommend a 6-point or 7-point quadrature rule for this element.
\end{remark}

\begin{remark}
Segal \cite{segal} explains 
for output purposes (printing, plotting etc.) the discontinuous pressures are averaged 
in vertices for all the adjoining elements. See also Fig. 7.3 of \cite{cuss86}.
\end{remark}

\begin{remark}
The simplest Crouzeix-Raviart element is the non-conforming linear triangle 
with constant pressure ($P_1\times P_0$) \cite{cuss86}. 
\end{remark}

It is worth noting that this element has more degrees of freedom  than the 
Taylor-Hood element for the same order of accuracy. However, since the 
bubble can be eliminated, one can design a modified version of this element.
\todo[inline]{Check Cuvelier book chapter 8 for modified element}


\begin{remark}
I have once asked the (main) author of MILAMIN why he chose this element, for 
example over the $P_2\times P_1$. His answer is as follows:
"Elements with continuous pressure  are incapable of converging in the Linf 
norm for mechanical problems exhibiting pressure jumps such as the inclusion-host setup. 
During my MSc and PhD I was focusing on sharp heterogeneities, so this is why I decided 
to choose $P_2^+\times P_{-1}$. 
You will see that it is also easy to invert the pressure mass matrix for such elements, 
which is really useful (both for the augmentation and preconditioning)."
\end{remark}

This element is used by Poliakov and Podlachikov \cite{popo92} to study the deformation of the surface above a rising diapir. Note that they actually use a "13 point integration formula (Hughes
1987) for calculation of the stiffness matrix was used in order
t o conserve detailed information from the marker field in
the coarse FEM mesh". 
It is also used in \cite{anmp15} in the context of a new free-surface stabilization scheme. 
It is the element used in LaCoDe \cite{demh19}.






%-----------------------------------------------------------------
\subsubsection{The Rannacher-Turek element - rotated $Q_1\times P_0$} \label{ss:RTq1p0}
\index{general}{$\tilde{Q}_1\times P_0$}
\index{general}{Korn's inequality}
\index{general}{Rannacher-Turek element}
\index{general}{Nonconforming element}

This element is the natural quadrilateral analogue
of the well-known triangular Stokes element of Crouzeix-Raviart \cite{crra73}.
This element is sometimes called $Q_1^{rot} \times Q_0$ or the Rannacher-Turek element 
\cite[Section 3.6.5]{john16}.
This rectangular nonconforming \cite{crfa89} element is termed the rotated $Q_1$ element 
because of the fact that $r^2-s^2$ can be generated from $rs$ (occurring in the bilinear $Q_1$ 
element) by a rotation of 45$\degree$ \cite[p93]{chen}.
The velocity approximation is achieved by rotated dim-linear functions that have 
continuous degrees of freedom on
the faces of the mesh cells as we have seen in Section~\ref{ss:rq1}.
This element was introduced in Rannacher \& Turek (1992) \cite{ratu92} 
has been proven to satisfy the inf-sup condition. It has been studied comprehensively in Schieweck 
(1997)\footnote{Habilitation thesis in German}, \cite{shzh06} and in Turek \cite{ture94,ture96}.
Superconvergence properties have also been reported \cite{misx06,misx07}.
It has been used in 2D \cite{maky17} and 3D \cite{klll96,gekm08} and forms the basis of the FeatFlow 
software\footnote{\url{http://www.featflow.de/en/index.html}}. 
It is used in the PhD thesis of Gastaldo \cite{gast07} and Ouazzi \cite{ouaz05}.
It has been 
successfully coupled to multigrid solvers \cite{chos98,tuos02}.
This element has been compared to the stabilised $Q_1\times P_0$ element \cite{lisi13}.
It is mentioned in \cite{hans11}

It essentially comes in two flavours, the Middle Point (MP) and the Mid Value (MV) one.

\begin{remark} 
John \cite{john16} explains that: "For the point-value-oriented non-conforming finite element spaces (MP), 
the value of the Dirichlet boundary
condition in the barycenter of the faces at the boundary is taken. Using the mean-
value-oriented spaces (MV), one computes the integrals of the boundary condition on
these faces and normalizes with the area of the faces to set the boundary values.
In the case of homogeneous Dirichlet boundary conditions, the boundary values
computed in both ways are zero."
\end{remark}

\begin{remark} 
John also makes a very important point: "There are also unmapped (non-parametric) versions of 
these finite element spaces, which define the polynomials directly on the mesh cell K. It is shown in Rannacher
and Turek (1992) \cite{ratu92} that these versions are inf-sup stable on more general meshes than
the mapped (parametric) version of the $Q_1^{rot}\times Q_0$ finite element, e.g., on strongly
nonuniform meshes. Considering all four types of $Q_1^{rot}\times Q_0$ finite elements, the
optimal order of convergence on perturbed meshes is achieved only by the mean-
value-oriented version of the unmapped $Q_1^{rot}\times Q_0$   finite element.
\end{remark}

Mahmood \etal \cite{maky17} mention a very important fact: "The chosen nonconforming element requires
additional stabilization for handling the deformation tensor formulation due to missing Korn's inequality 
\cite{horg95,knob00}.
To this end we employ the standard edge oriented stabilization \cite{tuos02,tuou07} in our simulations."
This is a rather unfortunate fact that although LBB stable this element needs an additional 
term in the weak form (see Turek \etal (2002) \cite{tuos02}) 
so as to suppress parasitic velocity modes when the div-grad formulation 
of the Stokes equation is used (as opposed to the Laplace formulation -- see \cite[Section 6.5.2]{dohu03}).

\Literature Sheen (2020) \cite{shee20}




%------------------------------------------------------------------
\subsubsection{The (stabilised) $P_1\times P_1$} \label{ss:P1P1stab}

Like its quadrilateral counterpart $Q_1\times Q_1$, the 
$P_1\times P_1$ pair is not stable and needs to be stabilised \cite{nosi98,tasu00}.
TerraNeo code uses stabilised with PSPG \cite{babd20}.





%----------------------------
\subsubsection{Other elements}

\begin{itemize}
\item $P_1\times P_0$: example 3.70 in \cite{john16}, also \cite{john98}. \index{general}{$P_1\times P_0$}
\item $Q_2\times P_0$: \index{general}{$Q_2 \times P_0$}: 
Quadratic velocities, constant pressure. The element satisfies the inf-sup condition, but the constant pressure assumption may require fine discretisation.

\item $Q_2\times Q_2$: This element is never used, probably because a) it is unstable, b) it is very costly. 
There is one reference to it in \cite{hufb86}.
\item $Q_1\times P_{-1}$ Bilinear velocities,  piecewise linear discontinuous polynomial pressure.
\item See Fortin \cite{fort81} for various stable low order elements other than the enriched $Q_1^+ \times P_0$
\item $Q_1\times Q_1$ + nonconforming null edge average \cite{fros07}
\item $P_3\times P_2$ mentioned in \cite{sten90}   \index{general}{$P_3\times P_2 element$}

\item $Q_2 \times Q_{-1}$: shown in Table~3.13-2 of Gresho \& Sani's book \cite{grsa}, 
and discussed in Section~3.13.6b of the book too. Not LBB stable, 1 chequerboard mode.
Used in \textcite{grsu02} (2002) and compared with $Q_1\times P_0$, $Q_2\times P_{-1}$ and 
$Q_2\times Q_1$ for thermal cavity problem with NS equations.

\end{itemize}

%.........................................................................
\subsubsection{A note about incompressibility and standard mixed methods}

What follows is nicely explained and demonstrated in John \etal \cite{jolm17}. In their 
example 1.1 they look at the velocity error of benchmark VJ2 (see Section~\ref{mms9}) 
which analytical solution is a zero velocity field. They show that for the MINI, 
Taylor-Hood and Crouzeix-Raviart triangular elements the velocity error grows 
with the magnitude of the rhs. They also make this statement:
\say{
there are important applications, e.g., natural
convection problems, where the pressure is larger than the velocity by orders
of magnitude. In such situations, one cannot expect to compute accurate
velocity fields with classical mixed methods, at least for low order methods.
}




